Quantitative processing of broadband data as implemented in a scientific splitbeam echosounder

### 2022-02-22 Paper/Manuscript.tex
Change to r_c for the range for sv in the MS

Based on the point scattering strength samples and the phase angle samples, single targets can be detected, and range and bearing to the single targets can be estimated. This is typically achieved through a single echo detection algorithm (SED). Here we will assume that the samples from the pulse compressed data $\ypc(\samplesymt)$ originating from single target already have been identified, noting that the number of samples after the detected target may be higher than those those before the peak to include scattering processes for beyond ideal point targets. The alongship angle $\along_n$, athwartship angle $\athw_n$ and range $r_n$ at the \emph{peak} power \prxe(\samplesymt) within the detected target is used as estimates for $\along_t$, $\athw_t$ and $r_t$, respectively. A simple pseudo SED algorithm is implemented in the code for illustrative purposes.
%
\begin{figure}
\includegraphics[width=16cm]{Fig_singleTarget}
\caption{\label{fi:SED} The \prxe(\samplesymt) power (upper) and split beam angles ($\along_t$ and $\athw_t$) (lower) for the single target. The vertical line corresponds to the range $r_t$ for the single target.}
\end{figure}

From the autocorrelation function of the matched filter signal, $\ymfauto(\samplesymt)$, the equivalent samples around the peak are extracted to create the reduced autocorrelation signal of the matched filter signal, $\ymfautored(\samplesymt)$. Depending on the target scattering characteristics and the distance to any adjacent single targets, the number of samples around the peak echo level in $\ypctarget(\samplesymt)$ that contain the majority of the echo energy can be more or less than the total number of samples around the peak of $\ymfauto(\samplesymt)$. If the number of samples around the target is more than the total number of samples around the peak of $\ymfauto(\samplesymt)$ all samples around the peak of $\ymfauto(\samplesymt)$ are used. If the number of samples around the target is less than the total number of samples around the peak of $\ymfauto(\samplesymt)$, this lower number is used to create a reduced autocorrelation signal, $\ymfautored(\samplesymt)$.

@@

Echoes from multiple scatterers can be quantified using volume backscattering strength, $\sv$, being the density of backscattering cross sections, and is given by:
%
\begin{equation}
\label{eq:sv}
\sv  =  10\log_{10}\frac{\sum\bs}{V}.
\end{equation}
%
where $V$ is the volume occupied by the scattering targets. The power-budget equation for multiple targets is then:
%
\begin{equation}
\label{eq:sv_f}
\sv(\freqsym) = 10\log_{10}(\prxevf(\freqsym)) + 20\log_{10}(\range) + 2\absorp(\freqsym)\range 
- 10\log_{10}\left( \frac{\ptxe \wlen^2 \cw \tslide \eqang(\freqsym) \gainzero^2(\freqsym)}{32\pi^2} \right), 
\end{equation}
%
where $\prxevf(\freqsym)$ is the received electric power in a matched load for the signal from a volume at frequency $\freqsym$, $\cw$ the sound speed, and $\tslide$ the duration of the time window, excluding the zero-padded portion if applied, used for evaluating the frequency spectrum. Note that $\range$ is the range to the centre of the range volume covered by $\tslide$, and the two-way equivalent beam angle, $\eqang$, is a function of frequency that is derived from an empirical estimate of $\eqang$ at the nominal frequency, $\fn$:
\begin{equation}
\label{eq:PsiFc}
\eqang(f) = \eqang(\fn)\left(\frac{\fn}{f}\right)^2.
\end{equation}

Volume backscattering samples compressed over the operational frequency band are estimated by applying Eq. \ref{eq:sv_f} to the received digitized power samples using the on-axis gain value with $f$ set to the centre frequency of the broadband pulse, $\fc$:
\begin{equation}
\label{eq:Sv}
\sv(\samplesymt)  =  10\log_{10}(\prxe(\samplesymt)) + 20\log_{10}(\range(\samplesymt)) + 2\absorp(\fc)\range(\samplesymt) 
- 10\log_{10}\left( \frac{\ptxe \wlen^2(\fc) \cw \teff \eqang(\fc) \gainzero^2(\fc)}{32\pi^2} \right),
\end{equation}
noting that $\sv(\samplesymt)$ is an average over frequency of all echoes received at sample $\samplesymt$. In this case, the time window, $\tslide$, is the effective pulse duration, $\teff$, resulting from pulse compression.

To estimate $\sv$ as a function of frequency a Fourier transform is used, repeatedly applied via a sliding window in range. However, the duration of this sliding window can be so long that the difference in spherical spreading loss compensation ($r^2$, implemented as the $20\log_{10}(\range)$ term in Eq. \ref{eq:Sv}) from the beginning of the window to the end can be significant, particularly for short range measurements. Thus, compensation for spreading loss is performed before applying the discrete Fourier transform. Absorption loss compensation is also range dependent (and frequency dependent), but since absorption loss compensation is insignificant for typical operating frequencies at short ranges and the difference in absorption loss compensation between the beginning and the end of the sliding window is insignificant at longer ranges, compensation for absorption loss is performed after applying the discrete Fourier transform.

Compensation of spherical spreading loss requires compensation of received power by a factor of $r^2$, and hence compensation of amplitude by a factor of $\range$:
%
\begin{equation}
\label{eq:spreadcomp}
\ypcspread(\samplesymt) = \ypc(\samplesymt)\range(\samplesymt).
\end{equation}
%
where $\ypcspread(\samplesymt)$ is the pulse compressed signal compensated for spherical spreading. A discrete Fourier transform is performed on the range compensated pulse compressed sample data using a normalized sliding Hanning window, $\hannw(\genidxsym)$. The duration, $\tslide$, of the sliding window is chosen as a compromise between along-beam range resolution and frequency resolution. We suggest that it be at least twice the pulse duration and for computational efficiency reasons should result in a number of samples, $\nw$, which is a power of 2.

The normalised Hanning window, $\hannwnorm$, is given by: 
%
\begin{equation}
\label{eq:hannw}
\hannwnorm(\genidxsym) = \frac{\hannw(\genidxsym)}{\left( \frac{||\hannw||_2}{\sqrt{\nw}} \right)}, i = \frac{-\nw}{2}, \ldots, \frac{\nw}{2}
\end{equation}
%
and the discrete Fourier transform of the windowed data, $\ypcvolumef(\samplesymf)$, is then obtained from:
%
\begin{equation}
\label{eq:FFT_volume}
\ypcvolumef(\samplesymf) = \dft_\ndft 
\left( \hannwnorm(\genidxsym) \left(\ypcspread (\genidxsym+\samplesymt) \left[ u(\genidxsym + \frac{\nw}{2}) - u(\genidxsym - \frac{\nw}{2}) \right] \right) \right),
\end{equation}
%
where $u(\genidxsym)$ is the step function and $\samplesymt$ is the sample data index for the centre of the sliding window. The discrete Fourier transform of the auto correlation function of the matched filter signal, $\ymfautof(\samplesymf)$, also needs to be evaluated at the same frequencies:
%
\begin{equation}
\label{eq:FFT_TX_Auto}
\ymfautof(\samplesymf)  =  \dft_\ndft (\ymfauto(\samplesymt)).
\end{equation}
%
The normalized discrete Fourier transform of the windowed data, $\ypcvolumenormf(\samplesymf)$, is then given by:
%
\begin{equation}
\label{eq:FFT_volume_norm}
\ypcvolumenormf(\samplesymf) = \frac{\ypcvolumef(\samplesymf)}{\ymfautof(\samplesymf)},
\end{equation}
%
and received power into a matched load, $\prxevf(\samplesymf)$, is estimated from:
%
\begin{equation}
\label{eq:prx_FFT_volume}
\prxevf(\samplesymf) = \nchannels \left( \frac{|\ypcvolumenormf(\samplesymf)|}{2\sqrt{2}} \right)^2 \left( \frac{|\zrxe+\ztde|}{|\zrxe|}\right)^2 \\
\frac{1}{|\ztde|}. % All channels (*4), matched load (/2), and effective values (/sqrt(2))
\end{equation}
%
Finally, the discretized estimate of $\sv(\freqsym)$, $\sv(\samplesymf)$, is given by:
%
\begin{equation}
\label{eq:Sv_FFT}
\sv(\samplesymf) = 10\log_{10}(\prxevf(\samplesymf)) + 2\absorp(\samplesymf) \range \\
- 10\log_{10}\left( \frac{\ptxe \wlen^2 \cw \tslide \eqang(\samplesymf) \gainzero^2(\samplesymf) }{32\pi^2} \right).
\end{equation}

>>>>

Based on the point scattering strength samples and the phase angle samples, single targets can be detected, and range and bearing to the single targets can be estimated. This is typically achieved through a single echo detection algorithm (SED). Here we will assume that the samples from the pulse compressed data $\ypc(\samplesymt)$ originating from single target already have been identified, noting that the number of samples after the detected target may be higher than those those before the peak to include scattering processes for beyond ideal point targets. The alongship angle $\along_n$, athwartship angle $\athw_n$ and range $r_n$ at the \emph{peak} power $\prxe(\samplesymt)$ within the detected target is used as estimates for $\along_t$, $\athw_t$ and $r_t$, respectively. A simple pseudo SED algorithm is implemented in the code for illustrative purposes.

From the autocorrelation function of the matched filter signal, $\ymfauto(\samplesymt)$, the equivalent number of samples around the peak are extracted to create the reduced autocorrelation signal of the matched filter signal, $\ymfautored(\samplesymt)$. Depending on the target scattering characteristics and the distance to any adjacent single targets, the number of samples around the peak echo level in $\ypctarget(\samplesymt)$ that contain the majority of the echo energy can be more or less than the total number of samples around the peak of $\ymfauto(\samplesymt)$. If the number of samples around the target is more than the total number of samples around the peak of $\ymfauto(\samplesymt)$ all samples around the peak of $\ymfauto(\samplesymt)$ are used. If the number of samples around the target is less than the total number of samples around the peak of $\ymfauto(\samplesymt)$, this lower number is used to create a reduced autocorrelation signal, $\ymfautored(\samplesymt)$.
%
\begin{figure}
\includegraphics[width=16cm]{Fig_singleTarget}
\caption{\label{fi:SED} The $\prxe(\samplesymt)$ power (upper) and split beam angles ($\along_t$ and $\athw_t$) (middle) for the single target. The vertical line corresponds to the range $r_t$ for the single target. The $\ymfautored(\samplesymt)$ (lower) is the autocorrelation function of the transmit signal reduced to the length of the target signal and aligned with the peak power of the target.}
\end{figure}

@@

Echoes from multiple scatterers can be quantified using volume backscattering strength, $\sv$, being the density of backscattering cross sections, and is given by:
%
\begin{equation}
\label{eq:sv}
\sv  =  10\log_{10}\frac{\sum\bs}{V}.
\end{equation}
%
where $V$ is the volume occupied by the scattering targets. The power-budget equation for multiple targets is then:
%
\begin{equation}
\label{eq:sv_f}
\sv(\freqsym) = 10\log_{10}(\prxevf(\freqsym)) + 20\log_{10}(\range_c) + 2\absorp(\freqsym)\range_r 
- 10\log_{10}\left( \frac{\ptxe \wlen^2 \cw \tslide \eqang(\freqsym) \gainzero^2(\freqsym)}{32\pi^2} \right), 
\end{equation}
%
where $\prxevf(\freqsym)$ is the received electric power in a matched load for the signal from a volume at frequency $\freqsym$, $\cw$ the sound speed, and $\tslide$ the duration of the time window, excluding the zero-padded portion if applied, used for evaluating the frequency spectrum. Note that $\range_c$ is the range to the centre of the range volume covered by $\tslide$, and the two-way equivalent beam angle, $\eqang$, is a function of frequency that is derived from an empirical estimate of $\eqang$ at the nominal frequency, $\fn$:
\begin{equation}
\label{eq:PsiFc}
\eqang(f) = \eqang(\fn)\left(\frac{\fn}{f}\right)^2.
\end{equation}

Volume backscattering samples compressed over the operational frequency band are estimated by applying Eq. \ref{eq:sv_f} to the received digitized power samples using the on-axis gain value with $f$ set to the centre frequency of the broadband pulse, $\fc$:
\begin{equation}
\label{eq:Sv}
\sv(\samplesymt)  =  10\log_{10}(\prxe(\samplesymt)) + 20\log_{10}(\range_c(\samplesymt)) + 2\absorp(\fc)\range_c(\samplesymt) 
- 10\log_{10}\left( \frac{\ptxe \wlen^2(\fc) \cw \teff \eqang(\fc) \gainzero^2(\fc)}{32\pi^2} \right),
\end{equation}
noting that $\sv(\samplesymt)$ is an average over frequency of all echoes received at sample $\samplesymt$. In this case, the time window, $\tslide$, is the effective pulse duration, $\teff$, resulting from pulse compression.

To estimate $\sv$ as a function of frequency a Fourier transform is used, repeatedly applied via a sliding window in range. However, the duration of this sliding window can be so long that the difference in spherical spreading loss compensation ($r_c^2$, implemented as the $20\log_{10}(\range_c)$ term in Eq. \ref{eq:Sv}) from the beginning of the window to the end can be significant, particularly for short range measurements. Thus, compensation for spreading loss is performed before applying the discrete Fourier transform. Absorption loss compensation is also range dependent (and frequency dependent), but since absorption loss compensation is insignificant for typical operating frequencies at short ranges and the difference in absorption loss compensation between the beginning and the end of the sliding window is insignificant at longer ranges, compensation for absorption loss is performed after applying the discrete Fourier transform.

Compensation of spherical spreading loss requires compensation of received power by a factor of $r_c^2$, and hence compensation of amplitude by a factor of $\range_c$:
%
\begin{equation}
\label{eq:spreadcomp}
\ypcspread(\samplesymt) = \ypc(\samplesymt)\range_c(\samplesymt).
\end{equation}
%
where $\ypcspread(\samplesymt)$ is the pulse compressed signal compensated for spherical spreading. A discrete Fourier transform is performed on the range compensated pulse compressed sample data using a normalized sliding Hanning window, $\hannw(\genidxsym)$. The duration, $\tslide$, of the sliding window is chosen as a compromise between along-beam range resolution and frequency resolution. We suggest that it be at least twice the pulse duration and for computational efficiency reasons should result in a number of samples, $\nw$, which is a power of 2.

The normalised Hanning window, $\hannwnorm$, is given by: 
%
\begin{equation}
\label{eq:hannw}
\hannwnorm(\genidxsym) = \frac{\hannw(\genidxsym)}{\left( \frac{||\hannw||_2}{\sqrt{\nw}} \right)}, i = \frac{-\nw}{2}, \ldots, \frac{\nw}{2}
\end{equation}
%
and the discrete Fourier transform of the windowed data, $\ypcvolumef(\samplesymf)$, is then obtained from:
%
\begin{equation}
\label{eq:FFT_volume}
\ypcvolumef(\samplesymf) = \dft_\ndft 
\left( \hannwnorm(\genidxsym) \left(\ypcspread (\genidxsym+\samplesymt) \left[ u(\genidxsym + \frac{\nw}{2}) - u(\genidxsym - \frac{\nw}{2}) \right] \right) \right),
\end{equation}
%
where $u(\genidxsym)$ is the step function and $\samplesymt$ is the sample data index for the centre of the sliding window. The discrete Fourier transform of the auto correlation function of the matched filter signal, $\ymfautof(\samplesymf)$, also needs to be evaluated at the same frequencies:
%
\begin{equation}
\label{eq:FFT_TX_Auto}
\ymfautof(\samplesymf)  =  \dft_\ndft (\ymfauto(\samplesymt)).
\end{equation}
%
The normalized discrete Fourier transform of the windowed data, $\ypcvolumenormf(\samplesymf)$, is then given by:
%
\begin{equation}
\label{eq:FFT_volume_norm}
\ypcvolumenormf(\samplesymf) = \frac{\ypcvolumef(\samplesymf)}{\ymfautof(\samplesymf)},
\end{equation}
%
and received power into a matched load, $\prxevf(\samplesymf)$, is estimated from:
%
\begin{equation}
\label{eq:prx_FFT_volume}
\prxevf(\samplesymf) = \nchannels \left( \frac{|\ypcvolumenormf(\samplesymf)|}{2\sqrt{2}} \right)^2 \left( \frac{|\zrxe+\ztde|}{|\zrxe|}\right)^2 \\
\frac{1}{|\ztde|}. % All channels (*4), matched load (/2), and effective values (/sqrt(2))
\end{equation}
%
Finally, the discretized estimate of $\sv(\freqsym)$, $\sv(\samplesymf)$, is given by:
%
\begin{equation}
\label{eq:Sv_FFT}
\sv(\samplesymf) = 10\log_{10}(\prxevf(\samplesymf)) + 2\absorp(\samplesymf) \range_c \\
- 10\log_{10}\left( \frac{\ptxe \wlen^2 \cw \tslide \eqang(\samplesymf) \gainzero^2(\samplesymf) }{32\pi^2} \right).
\end{equation}
